Computational biology in the 21st century: Algorithms that scale

### 2015-07-28 main.tex
More notes
new figure and minor tweaks

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
allowing scientists to also infer the expression of the gene to which it might translate).
However, the greatest volume of sequence data available are genomic in nature and 
come from next-generation sequencing (NGS) technologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.

Importantly to this Review, Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads are read concurrently and must be reassembled in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a book with all its pages torn out.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.

@@

\begin{figure}[htb!]
\centering
\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
\includegraphics[width=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
% YWY: 16S isn't well introduced
The difficulty is that the datasets required to shine any light on these 
questions, namely from ``shotgun'' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria based on 
their divergence with respect to a specific gene.
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that ``compressive omics''
will provide the ability for computation to scale sublinearly with massively growing data.

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as ``seeds'' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

In the area of metagenomic search,
the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to the
state-of-the-art DIAMOND~\cite{buchfink2014fast}, using it
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.

Compressive genomics \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well, and coined
``compressive omics''~\cite{yu2015entropy}.
One such example is chemogenomics.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.
Another example is 
esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

@@

The field of computational biology must continue to innovate, but also to 
incorporate the best ideas from other areas of computer science.
For example, the compressive-acceleration approach bears similarity to a metric 
ball tree, first described in the database community over twenty years 
ago~\cite{uhlmann1991satisfying};
however, the latter does not allow one to analyze performance guarantees, 
in terms of metric entropy and fractal dimension.
Other ideas from image processing, computational geometry, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
outside of biology are likely to bear fruit~\cite{indyk1998approximate}. 
In turn, it is also likely that algorithmic ideas developed within computational biology
will become useful in other fields experiencing a data deluge, such as astronomy
or social networks~\cite{stephens2015big}.

>>>>

% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
% YWY: My inclination is to write "(a) Moore's and (b) Kryder's laws". Possibly a matter of taste though, so do as you will.
\label{fig:growth}
\end{figure}

@@

% YWY: "Most" is an extremely strong claim. I believe you, but I could see people
% arguing (e.g. all of electronic health records could be considered "biological")
Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence and quantity, 
allowing scientists to also infer the expression of the gene to which it might translate).
However, the greatest volume of sequence data available are genomic in nature and 
come from next-generation sequencing (NGS) technologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.

Importantly to this Review, Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads are read concurrently and must be reassembled in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
This so-called `shotgun sequencing' is analogous to the problem of 
reconstructing a book with all its pages torn out.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.

@@

% YWY: Take spell-check lines out of figure labels! (bp, Samtools)
\begin{figure}[htb!]
\centering
\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
\includegraphics[width=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

@@

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
% YWY: 16S isn't well introduced
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than the 16S metagenomic studies that merely classify the bacteria 
based on 
their divergence with respect to a specific gene (namely, the 16S ribosomal 
subunit).
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} had been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

@@

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
Due to the complexity of protein structures, these programs generally take 
significant amounts of time,
especially to align multiple structures.
For example, the DALI webserver can take as much as an hour to return results
for a single query.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments
by instead comparing the `bag-of-words' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

@@

Techniques for reference-based read mapping typically rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} mapper further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation of the 
reference genome, in order to prune the search space 
when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} and mrsFAST \cite{hach2014mrsfast} use an `approximate seed' approach to indexing
the space of possible matches, likewise pruning the search space; however, the
bulk of its runtime is spent on the extend phase.
State-of-the-art mapper mrsFAST-Ultra achieves improvements in efficiency based on machine architecture rather than leveraging redundancy in the data itself with near-perfect sensitivity, but
only for the case where there are no insertions and deletions (indels) \cite{hach2014mrsfast}.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
`shaped seeds' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that `compressive omics'
will provide the ability for computation to scale sublinearly with massively growing data.

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each one to its neighbors within
a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

In the area of metagenomic search,
the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to the
state-of-the-art DIAMOND~\cite{buchfink2014fast}, using it
for its `coarse search' phase and a user's choice of DIAMOND or BLASTX for its
`fine search' phase; MICA demonstrates nearly order-of-magnitude run-time gains over the highly-optimized DIAMOND, comparable to that of caBLASTP over BLASTP.

Compressive genomics \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well, and coined
`compressive omics'~\cite{yu2015entropy}.
One such example is chemogenomics.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.
Another example is 
esFragBag~\cite{yu2015entropy}, which clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a `boosting' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

@@

The field of computational biology must continue to innovate, but also to 
incorporate the best ideas from other areas of computer science.
For example, the compressive-acceleration approach bears similarity to a metric 
ball tree, first described in the database community over twenty years 
ago~\cite{uhlmann1991satisfying};
however, the latter does not allow one to analyze performance guarantees, 
in terms of metric entropy and fractal dimension.
Other ideas from image processing, computational geometry~\cite{indyk1998approximate}, sublinear-time algorithms~\cite{rubinfeld2011sublinear}, and other areas 
outside of biology are likely to bear fruit. 
In turn, it is also likely that algorithmic ideas developed within computational biology
will become useful in other fields experiencing a data deluge, such as astronomy
or social networks~\cite{stephens2015big}.
